---
title: "ChIP-seq Visualization with trackplot"
subtitle: "ENCODE H1 hESC Data — OCT4/NANOG Locus & Genome-Wide Profiles"
author: "Danya Hassan"
date: today
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    theme: flatly
    highlight-style: tango
    code-fold: show
    self-contained: true
execute:
  warning: false
  message: false
---

---

## 1. Overview

This notebook demonstrates genome-scale ChIP-seq visualization using the
[**trackplot**](https://github.com/PoisonAlien/trackplot) R package, applied
to publicly available ENCODE data from H1 human embryonic stem cells (hESCs).

Two complementary visualization types are covered:

- **Locus (IGV-style) track plots** — signal tracks, peak annotations, and
  ChromHMM chromatin state tracks at the OCT4/POU5F1 locus
- **Metagene profile plots** — genome-wide average signal around RefSeq
  transcription start sites (TSS)

**Data source:** [ENCODE Project](https://www.encodeproject.org/) — H1 hESC
ChIP-seq experiments for OCT4, NANOG, H3K27ac, H3K4me1, H3K4me3, CTCF,
H2A.Z, and RNA Pol II.

---

## 2. Environment Setup

### 2.1 Verify bwtool is Accessible

trackplot calls `bwtool` as a system command to extract signal from bigWig
files. This block confirms it is on the R session's PATH before proceeding.

```{r verify-bwtool}
# If running outside Docker, manually add bwtool to R's PATH:
# Sys.setenv(PATH = paste(Sys.getenv("PATH"), "/home/<user>/bin", sep = ":"))

bwtool_path <- Sys.which("bwtool")

if (nchar(bwtool_path) == 0) {
  stop(
    "bwtool not found on PATH.\n",
    "Add it with: Sys.setenv(PATH = paste(Sys.getenv('PATH'), '/path/to/bin', sep = ':'))\n",
    "See README for full installation instructions."
  )
} else {
  cat("bwtool found at:", bwtool_path, "\n")
  system("bwtool 2>&1 | head -2")
}
```

### 2.2 Load Libraries

```{r libraries}
library(data.table)
library(ggplot2)
library(dplyr)
library(tidyr)

# Install trackplot if not already available
if (!requireNamespace("trackplot", quietly = TRUE)) {
  remotes::install_github("poisonalien/trackplot")
}
library(trackplot)
```

---

## 3. Configuration

```{r config}
# ── Edit paths here ──────────────────────────────────────────────────────────
WORKDIR <- "~/trackplot_tutorial"   # directory containing bigWig and bed files
BUILD   <- "hg19"                   # genome assembly used for ENCODE H1 data
# ─────────────────────────────────────────────────────────────────────────────

setwd(WORKDIR)
```

---

## 4. ENCODE Data Used

The following ChIP-seq experiments from the H1 hESC ENCODE tier-1 cell line
are used in this analysis. BigWig signal files and narrowPeak BED files were
downloaded directly from the ENCODE portal.

| Mark / Factor | ENCODE Accession | Type |
|--------------|-----------------|------|
| OCT4 (POU5F1) | ENCFF465CMT | BigWig |
| NANOG | ENCFF996VEE | BigWig |
| H3K27ac | — | BigWig |
| H3K4me1 | — | BigWig |
| H3K4me3 | — | BigWig |
| CTCF | H1_Ctcf.bed | NarrowPeak |
| H2A.Z | H1_H2az.bed | NarrowPeak |
| RNA Pol II | H1_Pol2.bed | NarrowPeak |

---

## 5. Locus Track Plots (IGV-Style)

### 5.1 Load BigWig Files

```{r load-bigwigs}
# List all bigWig files in the working directory
bigwig_files <- list.files(pattern = "\\.bigWig$|\\.bw$")
cat("BigWig files found:", length(bigwig_files), "\n")
print(bigwig_files)

# Build colData table linking files to genome assembly
bigWigs <- read_coldata(bws = bigwig_files, build = BUILD)
head(bigWigs)
```

### 5.2 Extract Signal at OCT4/POU5F1 Locus

trackplot can accept either a genomic coordinate range or a gene name.
Both approaches are shown below — they produce identical output.

```{r extract-signal}
oct4_loci <- "chr6:31125776-31144789"

# Option A: extract by genomic coordinates
t_coords <- track_extract(colData = bigWigs, loci = oct4_loci)

# Option B: extract by gene name (fetches coordinates from UCSC automatically)
t_gene <- track_extract(colData = bigWigs, gene = "POU5F1")

# Use gene-based extraction for all downstream plots
t <- t_gene
```

### 5.3 Basic Track Plot

```{r basic-trackplot}
track_plot(summary_list = t)
```

### 5.4 Custom Colors + Chromosome Ideogram

```{r trackplot-ideogram}
track_cols <- c("#d35400", "#27ae60")

track_plot(
  summary_list  = t,
  col           = track_cols,
  show_ideogram = TRUE
)
```

### 5.5 Overlay All Tracks into a Single Track

Overlaying tracks is useful for directly comparing signal levels between
conditions or replicates at the same locus.

```{r trackplot-overlay}
track_plot(
  summary_list  = t,
  col           = track_cols,
  show_ideogram = TRUE,
  track_overlay = TRUE
)
```

### 5.6 Multi-Gene View with Auto-Scaled Groups

Display multiple genes in the same window with independent y-axis scaling
per group, which is helpful when comparing factors with different signal
dynamic ranges.

```{r trackplot-multigene}
track_plot(
  summary_list   = t,
  col            = track_cols,
  show_ideogram  = TRUE,
  genename       = c("POU5F1", "TCF19"),
  groupAutoScale = TRUE,
  y_min          = 0
)
```

---

## 6. NarrowPeak / BroadPeak Tracks

Peak BED files can be loaded alongside bigWig signal tracks to show called
peaks directly on the locus plot.

```{r load-peaks}
narrowPeak_files <- c(
  "H1_Ctcf.bed",
  "H1_H2az.bed",
  "H1_k27ac.bed",
  "H1_k4me1.bed",
  "H1_k4me3.bed",
  "H1_Nanog.bed",
  "H1_Oct4.bed",
  "H1_Pol2.bed"
)

# Confirm files exist before loading
narrowPeak_files <- narrowPeak_files[file.exists(narrowPeak_files)]
cat("Peak files found:", length(narrowPeak_files), "\n")

narrowPeaks <- read_coldata(
  narrowPeak_files,
  build      = "hg19",
  input_type = "peak"
)
```

```{r extract-peaks}
# Wider genomic window to show peak context across the OCT4/NANOG region
oct4_region <- "chr6:30818383-31452182"   # ~633 kb window

narrowPeaks_track <- track_extract(
  colData = narrowPeaks,
  loci    = oct4_region
)
```

```{r plot-peaks}
# Overlay ChromHMM chromatin state tracks from UCSC alongside peak tracks
track_plot(
  summary_list = narrowPeaks_track,
  show_ideogram = TRUE,
  ucscChromHMM  = c(
    "wgEncodeBroadHmmH1hescHMM",   # H1 hESC chromatin states
    "wgEncodeBroadHmmNhlfHMM"      # NHLF (lung fibroblast) for comparison
  )
)
```

---

## 7. Metagene Profile Plots (Genome-Wide)

Profile plots show the average ChIP-seq signal across all RefSeq transcripts,
centered on the transcription start site (TSS). This provides a genome-wide
view of how a factor distributes relative to gene bodies.

### 7.1 Load BigWigs for Profile Analysis

```{r load-profile-bws}
profile_bws <- c("ENCFF465CMT.bigWig", "ENCFF996VEE.bigWig")

bws <- read_coldata(
  bws          = profile_bws,
  sample_names = c("OCT4_rep1", "OCT4_rep2"),
  build        = "hg38"
)
```

### 7.2 Extract Signal Around TSS

Signal is extracted in a window extending 1,500 bp upstream and downstream
of the TSS for all RefSeq transcripts.

```{r profile-extract}
pe_refseq <- profile_extract(
  colData       = bws,
  ucsc_assembly = TRUE,
  startFrom     = "start",   # anchor to TSS
  up            = 1500,      # bp upstream of TSS
  down          = 1500       # bp downstream of TSS
)
```

### 7.3 Summarize and Plot

```{r profile-plot}
ps_refseq <- profile_summarize(sig_list = pe_refseq)
profile_plot(ps_refseq)
```

---

## 8. Generic Usage Template

This section provides a minimal template for running trackplot on **any
bigWig dataset** — replace file names, genome build, and locus as needed.

```{r generic-template, eval=FALSE}
# ── 1. Point to your bigWig files ────────────────────────────────────────────
my_bws <- list.files(path = "path/to/bigwigs/", pattern = "\\.bw$", full.names = TRUE)

# ── 2. Build colData (swap build for hg19, hg38, mm10, etc.) ─────────────────
my_coldata <- read_coldata(bws = my_bws, build = "hg38")

# ── 3. Extract signal at a locus or gene ─────────────────────────────────────
my_tracks <- track_extract(colData = my_coldata, gene = "GAPDH")
# OR by coordinates:
# my_tracks <- track_extract(colData = my_coldata, loci = "chr1:1000000-1050000")

# ── 4. Plot ───────────────────────────────────────────────────────────────────
track_plot(
  summary_list  = my_tracks,
  show_ideogram = TRUE,
  col           = c("#e74c3c", "#3498db", "#2ecc71")   # one color per track
)

# ── 5. Metagene profile (optional) ───────────────────────────────────────────
pe  <- profile_extract(colData = my_coldata, ucsc_assembly = TRUE,
                        startFrom = "start", up = 2000, down = 2000)
ps  <- profile_summarize(sig_list = pe)
profile_plot(ps)
```

---

## 9. Troubleshooting

| Problem | Solution |
|---------|----------|
| `bwtool not found` | `Sys.setenv(PATH = paste(Sys.getenv("PATH"), "~/bin", sep=":"))` |
| `mysql: command not found` | Install MySQL client: `conda install -c anaconda mysql` |
| `Error in read_coldata`: file not found | Check working directory with `getwd()` and confirm file names match exactly |
| UCSC queries timing out | Try again — UCSC browser has rate limits; or download annotation files locally |
| ChromHMM track names not found | Check available track names at https://genome.ucsc.edu/cgi-bin/hgTrackDb |

---

## 10. Session Info

```{r session-info}
sessionInfo()
```
